---
title: "Reproducible Research - Peer-graded Assignment: Course Project 1"
author: "pduchesne"
date: "25 February 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide")
```
### Load libraries

Load DLPYR, RColorBrewer & Hmisc libraries

```{r, echo = TRUE, message = FALSE}
library(Hmisc)
library(dplyr)
library(RColorBrewer)

```

Load source file and extract

```{r ETL,  echo=TRUE}
# verify if activity.zip exists in the working directory, if not download file.  
if(!file.exists("./activity.zip")){
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileUrl, destfile="./activity.zip")
}

# extract file if "/course project 1/activity" directory is not present
if(!file.exists("./activity")){
        unzip("./activity.zip", files = NULL, list = FALSE, overwrite = TRUE,
              junkpaths = FALSE, exdir = "./activity", unzip = "internal",
              setTimes = FALSE)
}
```

## What is mean total number of steps taken per day?
Load file into data frame.
```{r t_load, results='asis', echo=TRUE}
df_activity <- read.csv("./activity/activity.csv", na.strings="NA", stringsAsFactors=FALSE)
df_activity$date<-as.Date.character(df_activity$date,format = "%Y-%m-%d")
```

1. Calculate the total number of steps taken per day.
```{r t_steps, results='asis', echo=TRUE}
df_daily <- aggregate(df_activity$steps, by=list(df_activity$date), FUN=sum, na.rm=TRUE)
names(df_daily)<-c("date","steps")
```

2. Make a histogram of the total number of steps taken each day
## Histogram of the total number of steps taken each day
```{r hist, echo=TRUE}
hist(df_daily$steps,col=brewer.pal(5,"Set3"),
main="Histogram of the total number of steps taken each day",
xlab="number of steps taken each day"
)
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r M_M, results='asis', echo=TRUE}
df_M_M <- data.frame(Mean = mean(df_daily$steps, na.rm=TRUE), Median = median(df_daily$steps, na.rm=TRUE))
knitr::kable(df_M_M)
```

## What is the average daily activity pattern?

1.  Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r avg_steps, echo=TRUE}
df_interval_avg <- df_activity %>% group_by(interval) %>% summarize(mean_steps = mean(steps, na.rm = TRUE))
plot(df_interval_avg$interval,df_interval_avg$mean_steps, type="l", xlab="Intervals",ylab = "Average Steps")
```

2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max_avg_steps , results='asis', echo=TRUE}
mx <-max(df_interval_avg$mean_steps)
five_minute_interval <- df_interval_avg %>% select(interval) %>% filter(df_interval_avg$mean_steps == mx)
knitr::kable(five_minute_interval)
```

## Imputing missing values
1. The total number of rows with NA:

```{r Nas, results='asis', echo=TRUE}
df_missing<-sum(is.na(df_activity$steps))
names(df_missing)<-"Total : "
knitr::kable(df_missing)
```


2.  Strategy for filling in all of the missing values in the dataset: use **Hmisc** package

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r wo_na , results='asis', echo=TRUE}
df_activity_wo_na <- df_activity
df_activity_wo_na$steps <- with(df_activity, impute(steps, mean))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

## Histogram of the total number of steps taken each day (without NA)
```{r hist_wo_na, echo=TRUE}
df_daily_wo_na <- aggregate(df_activity_wo_na$steps, by=list(df_activity_wo_na$date), FUN=sum, na.rm=TRUE)
names(df_daily_wo_na)<-c("date","steps")

hist(df_daily_wo_na$steps,col=brewer.pal(5,"Set3"),
main="Histogram of the total number of steps taken each day (without NA)",
xlab="number of steps taken each day")
```

and calculate and report the mean and median of the total number of steps taken per day
```{r M_M_wo_na, results='asis', echo=TRUE}
df_M_M_wo_na <- data.frame(Mean = mean(df_daily_wo_na$steps, na.rm=TRUE), Median = median(df_daily_wo_na$steps, na.rm=TRUE))
knitr::kable(df_M_M_wo_na)
```

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

### Answer: 
#### Yes, as the frequency for the central bar of the histogram (that holding the replacement for NA with the average value 10766.19) has increased from 25 to 35. The bar representing 0 to 5000 has decreased accordingly.

## Are there differences in activity patterns between weekdays and weekends?

#### Create a new factor variable in the dataset with two levels - "weekday" and "weekend"

```{r weekend, results='hide', echo=TRUE}
df_activity_wo_na$day_type[weekdays(df_activity_wo_na$date, abbreviate = TRUE) %in% c("Sat","Sun")] <- "Weekend"
df_activity_wo_na$day_type[is.na(df_activity_wo_na$day_type)]<-"Weekday"
df_activity_wo_na$day_type<- as.factor(df_activity_wo_na$day_type)

df_interval_avg_wo_na <- df_activity_wo_na %>% group_by(interval,day_type) %>% summarize(mean_steps = mean(steps, na.rm = TRUE))

```

#### Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r xyplot, results='asis', echo=TRUE}
xyplot(df_interval_avg_wo_na$mean_steps ~ df_interval_avg_wo_na$interval|df_interval_avg_wo_na$day_type,
            main="Differences in activity patterns between weekdays and weekends",
            xlab="Intervals",ylab="Number of steps",type = "l",
            layout=c(1,2))
```